---
title: "1_2_dataexp_dimensionalityred"
output: html_document
---

# Excersize 1:
## Report sample count by transcriptome subtype (paper_Transcriptome.Subtype): Classical (CL), Mesenchymal (ME), Neural (NE), Proneural (PN).
```{r}
#load important libraries
library(here)
library(ggplot2)
library(patchwork)
library(dplyr)
```

At first we load the data
```{r}
load(here("data/GBM_data.RData"))
```
Next we can report sample counts by transcriptome subtype
```{r}
table(gbm_anot_clean$paper_Transcriptome.Subtype)
```

##  Are there recurrent tumor samples (shortLetterCode = TR)? How will you treat them?
```{r}
any(gbm_anot_clean == "TR", na.rm = T)
```
# Excersize 2:
##  PCA (PC1 vs. PC2, PC1 vs. PC3): color by transcriptome subtype and by IDH status (paper_IDH.status).

At first we want to transform our matrix and scale the data so that the gene expression for each gene is normally distributed with mean 0 and variance of 1
```{r}

# Remove gene column and set rownames to gene names
t_gbm_exp_clean <- gbm_exp_clean
t_gbm_exp_clean <- t_gbm_exp_clean[,-1]
t_gbm_exp_clean <- t(t_gbm_exp_clean)
colnames(t_gbm_exp_clean) <- gbm_exp_clean$gene

# Now normalize the gene expression levels using Z-transformation
norm_t_gbm_exp_clean <- scale(t_gbm_exp_clean)

# Next we compute the principle compontent analysis
pca_gbm_exp <- prcomp(norm_t_gbm_exp_clean, center = F, scale = F)
pca_gbm_scores <- as.data.frame(pca_gbm_exp$x)
pca_gbm_scores$barcode = rownames(pca_gbm_scores)
plot_pca_gbm_df <- merge(pca_gbm_scores, gbm_anot_clean[, c("paper_Transcriptome.Subtype", "paper_IDH.status", "barcode")], by = "barcode")

#Compute Variance explained by each PC
pca_gbm_var_explained <- pca_gbm_exp$sdev^2 / sum(pca_gbm_exp$sdev^2)

```
We also want to select the loadings of the PCs so we can visualize then in the plots as well
```{r}
pca_gbm_loadings <- as.data.frame(pca_gbm_exp$rotation)
# Add a row equivalent to the rownames (genes) to the data.frame
pca_gbm_loadings$genes <- rownames(pca_gbm_loadings)
# Next we select only the genes which have the highest absolute loadings 
pca_gbm_p1_top <- pca_gbm_loadings[
  order(abs(pca_gbm_loadings$PC1), decreasing = TRUE),
][1:10, c("PC1", "genes")]

pca_gbm_p2_top <- pca_gbm_loadings[
  order(abs(pca_gbm_loadings$PC2), decreasing = TRUE),
][1:10, c("PC2", "genes")]

pca_gbm_p3_top <- pca_gbm_loadings[
  order(abs(pca_gbm_loadings$PC3), decreasing = TRUE),
][1:10,c("PC3", "genes")]
```


```{r}
#Next we plot PC1 against PC2 and PC3 coloring each element according to transcriptome subtype and IDH status

p1 <- ggplot(plot_pca_gbm_df, 
             aes(x = PC1, y = PC2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PC1 vs PC2", color = "Subtype")

p2 <- ggplot(plot_pca_gbm_df, 
             aes(x = PC1, y = PC3, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PC1 vs PC3", color = "Subtype")

p3 <- ggplot(plot_pca_gbm_df, 
             aes(x = PC1, y = PC2, color = paper_IDH.status)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PC1 vs PC2", color = "IDH status")

p4 <- ggplot(plot_pca_gbm_df, 
             aes(x = PC1, y = PC3, color = paper_IDH.status)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PC1 vs PC3", color = "IDH status")


(p1 | p2) /
(p3 | p4)
```
We can see that the subtypes are especially well separated in the PC1 vs PC2 plot along the PC2 axis. 
```{r}
#Show the top loadings of the PCs
p_pc1 <- ggplot(pca_gbm_p1_top, aes(x = genes, y = PC1)) +
  geom_col(color = "white", fill = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top loadings PC1", x = "Gene", y = "Loading")

p_pc2 <- ggplot(pca_gbm_p2_top, aes(x = genes, y = PC2)) +
  geom_col(color = "white", fill = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top loadings PC2", x = "Gene", y = "Loading")

p_pc3 <- ggplot(pca_gbm_p3_top, aes(x = genes, y = PC3)) +
  geom_col(color = "white", fill = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top loadings PC3", x = "Gene", y = "Loading")

(p_pc1 | p_pc2) / p_pc3
```
## t-SNE: do the four subtypes separate clearly? Which subtypes are most similar?
Next we want to perform t-SNE and visualize the data in a plot.
```{r}
# t-SNE is performed on the PCA data meaning that instead of all genes we enter a subset of PCs so that the distance between each point (patient) is more meaningful. To select how many PCs we choose we can visualize the cumulative variance explained by the PCs first.
plot(cumsum(pca_gbm_var_explained),  xlab = "Number of PCs",
     ylab = "Cumulative variance explained")
# Next we define how many PCs we want to include in the t-SNE analysis
n_pcs_tsne <- 30
tsne_input <- pca_gbm_scores[, 1:n_pcs_tsne]
# We also need to install and load Rtsne 
library("Rtsne")
# next we run the analysis with a seed for reproducibility
set.seed(42)

tsne_result_gbm <- Rtsne(
  tsne_input,
  dims = 2,              # for the 2 D representation
  perplexity = 10,       # adjust based on sample size (5–50 typical)
  verbose = TRUE,
  max_iter = 1000,
  check_duplicates = FALSE
)

# Now we can extract the 2-D coordinates from the results
tsne_df <- data.frame(
  TSNE1 = tsne_result_gbm$Y[,1],
  TSNE2 = tsne_result_gbm$Y[,2]
)
tsne_df$barcode = rownames(pca_gbm_scores)
# And we also add the subtype from the annotation data
tsne_df_anot = merge(tsne_df, gbm_anot_clean, by = "barcode")

# Next we need to plot the data and color it according to their respective subtype
ggplot(tsne_df_anot, aes(x = TSNE1, y = TSNE2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(x = "t-SNE 1", y = "t-SNE 2", color = "Subtype")
```
## UMAP on expression and on PCs: compare embeddings. Does the number of PCs affect subtype separation?
```{r}
# First we need to load the uwot package for UMAP 
library("uwot")

# At first we will compute the UMAP on the scaled expression data we also used for the pca.
set.seed(42)
umap_exp_gbm <- umap(norm_t_gbm_exp_clean)
# We will store the values for the data in a data.frame and also include the annotations
umap_exp_gbm_df <- as.data.frame(umap_exp_gbm)
umap_exp_gbm_df$barcode <- rownames(umap_exp_gbm_df)
umap_exp_gbm_df_anot <- merge(umap_exp_gbm_df, gbm_anot_clean, by = "barcode")

# Next we will perform the same thing on the PCA data with different numbers of PCs
pc_sets_umap <- list(
  PCs_10 = pca_gbm_scores[,1:10],
  PCs_30 = pca_gbm_scores[,1:30],
  PCs_50 = pca_gbm_scores[,1:50]
)
umap_pca_results <- lapply(pc_sets_umap, function(x) {
  umap(x)
})
umap_pca_results_df <- data.frame(PCA_10 = umap_pca_results$PCs_10[,1:2],
                                  PCA_30 = umap_pca_results$PCs_30[,1:2],
                                  PCA_50 = umap_pca_results$PCs_50[,1:2]
                                  )
umap_pca_results_df$barcode <- rownames(umap_pca_results_df)
umap_pca_results_df_anot <- merge(umap_pca_results_df, gbm_anot_clean, by = "barcode")
```
Next we want to plot the data
```{r}
umap_p1 <- ggplot(data = umap_exp_gbm_df_anot, aes(x = V1, y = V2, color = paper_Transcriptome.Subtype)) +
  geom_point(alpha = 0.8) +
  theme_minimal() + 
  labs(x = "UMAP 1", y = "UMAP 2", color = "Subtype")

umap_p2 <- ggplot(data = umap_pca_results_df_anot, aes(x = PCA_10.1, y = PCA_10.2, color = paper_Transcriptome.Subtype)) + 
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "UMAP 1 from 10 PCs", y = "UMAP 2 from 10 PCs", color = "Subtype")

umap_p3 <- ggplot(data = umap_pca_results_df_anot, aes(x = PCA_30.1, y = PCA_30.2, color = paper_Transcriptome.Subtype)) + 
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "UMAP 1 from 30 PCs", y = "UMAP 2 from 30 PCs", color = "Subtype")

umap_p4 <- ggplot(data = umap_pca_results_df_anot, aes(x = PCA_50.1, y = PCA_50.2, color = paper_Transcriptome.Subtype)) + 
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "UMAP 1 from 50 PCs", y = "UMAP 2 from 50 PCs", color = "Subtype")

(umap_p1 | umap_p2) / (umap_p3 | umap_p4)
```
```{r}
umap_p1.1 <- ggplot(data = umap_exp_gbm_df_anot, aes(x = V1, y = V2, color = paper_Pan.Glioma.RNA.Expression.Cluster)) +
  geom_point(alpha = 0.8) +
  theme_minimal() + 
  labs(x = "UMAP 1", y = "UMAP 2", color = "Subtype")

umap_p2.1 <- ggplot(data = umap_pca_results_df_anot, aes(x = PCA_10.1, y = PCA_10.2, color = paper_Pan.Glioma.RNA.Expression.Cluster)) + 
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "UMAP 1 from 10 PCs", y = "UMAP 2 from 10 PCs", color = "Subtype")

umap_p3.1 <- ggplot(data = umap_pca_results_df_anot, aes(x = PCA_30.1, y = PCA_30.2, color = paper_Pan.Glioma.RNA.Expression.Cluster)) + 
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "UMAP 1 from 30 PCs", y = "UMAP 2 from 30 PCs", color = "Subtype")

umap_p4.1 <- ggplot(data = umap_pca_results_df_anot, aes(x = PCA_50.1, y = PCA_50.2, color = paper_Pan.Glioma.RNA.Expression.Cluster)) + 
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "UMAP 1 from 50 PCs", y = "UMAP 2 from 50 PCs", color = "Subtype")

(umap_p1.1 | umap_p2.1) / (umap_p3.1 | umap_p4.1)
```


