---
title: "1_2_dataexp_dimensionalityred"
output: html_document
---

# Excersize 1:
## Report sample count by transcriptome subtype (paper_Transcriptome.Subtype): Classical (CL), Mesenchymal (ME), Neural (NE), Proneural (PN).

At first we load the data
```{r}
#load(here(data/GBM_data.RData))
```
Next we can report sample counts by transcriptome subtype
```{r}
table(gbm_anot_clean$paper_Transcriptome.Subtype)
```

##  Are there recurrent tumor samples (shortLetterCode = TR)? How will you treat them?
```{r}
any(gbm_anot_clean == "TR", na.rm = T)
```
# Excersize 2:
##  PCA (PC1 vs. PC2, PC1 vs. PC3): color by transcriptome subtype and by IDH status (paper_IDH.status).

At first we want to transform our matrix and scale the data so that the gene expression for each gene is normally distributed with mean 0 and variance of 1
```{r}
library(dplyr)
# Remove gene column and set rownames to gene names
t_gbm_exp_clean <- gbm_exp_clean
t_gbm_exp_clean <- t_gbm_exp_clean[,-1]
t_gbm_exp_clean <- t(t_gbm_exp_clean)
colnames(t_gbm_exp_clean) <- gbm_exp_clean$gene

# Now normalize the gene expression levels using Z-transformation
norm_t_gbm_exp_clean <- scale(t_gbm_exp_clean)

# Next we compute the principle compontent analysis
pca_gbm_exp <- prcomp(norm_t_gbm_exp_clean)
pca_gbm_scores <- as.data.frame(pca_gbm_exp$x)
pca_gbm_scores$barcode = rownames(pca_gbm_scores)
plot_pca_gbm_df <- merge(pca_gbm_scores, gbm_anot_clean[, c("paper_Transcriptome.Subtype", "paper_IDH.status", "barcode")], by = "barcode")

#Compute Variance explained by each PC
pca_gbm_var_explained <- pca_gbm_exp$sdev^2 / sum(pca_gbm_exp$sdev^2)

#Next we plot PC1 against PC2 and PC3 coloring each element according to transcriptome subtype and IDH status
library(ggplot2)

p1 <- ggplot(plot_pca_gbm_df, 
             aes(x = PC1, y = PC2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PC1 vs PC2", color = "Subtype")

p2 <- ggplot(plot_pca_gbm_df, 
             aes(x = PC1, y = PC3, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PC1 vs PC3", color = "Subtype")

p3 <- ggplot(plot_pca_gbm_df, 
             aes(x = PC1, y = PC2, color = paper_IDH.status)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PC1 vs PC2", color = "IDH status")

p4 <- ggplot(plot_pca_gbm_df, 
             aes(x = PC1, y = PC3, color = paper_IDH.status)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PC1 vs PC3", color = "IDH status")

library(patchwork)

(p1 | p2) /
(p3 | p4)
```

