---
title: "3_4_expressiondist_differentialexp"
output: html_document
---
```{r}
#load packages
library(dplyr)
library(ggplot2)
library(here)
library(edgeR)
library(patchwork)
library(pheatmap)
```
Load the data
```{r}
load(here("data/GBM_data.RData"))
```
Data cleaning
```{r}
# Remove gene column and set rownames to gene names
t_gbm_exp_clean <- gbm_exp_clean
t_gbm_exp_clean <- t_gbm_exp_clean[,-1]
t_gbm_exp_clean <- t(t_gbm_exp_clean)
colnames(t_gbm_exp_clean) <- gbm_exp_clean$gene
```
# Exercise 3:
## Are logCPM values approximately normal? Compare Q-Q plots across subtypes.
```{r}
t_gbm_exp_clean <- as.data.frame(t_gbm_exp_clean)
t_gbm_exp_clean$barcode <- rownames(t_gbm_exp_clean)

annot_t_gbm_exp_clean <- merge(t_gbm_exp_clean, gbm_anot_clean[, c("paper_Transcriptome.Subtype", "paper_IDH.status", "barcode")], by = "barcode")
```
Compute QQ-Plot on distribution of expression values within samples, grouped by subtype
```{r}
expr <- annot_t_gbm_exp_clean[, !(names(annot_t_gbm_exp_clean) %in% 
  c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status"))]

subtypes <- annot_t_gbm_exp_clean$paper_Transcriptome.Subtype

values <- as.numeric(as.matrix(expr))
Subtype <- rep(subtypes, each = ncol(expr))  

plot_df <- data.frame(value = values, Subtype = Subtype)


```
## plot side by side (takes a while to load)!
```{r}
ggplot(plot_df, aes(sample = value)) +
  stat_qq(size = 0.5) +
  stat_qq_line(color = "red") +
  facet_wrap(~ Subtype) +
  theme_minimal()
```
## CPM values appear to be roughly normaly distributed
##Question 2:Does normalization method (quantile vs. TMM vs. none) affect the separation observed in PCA?
```{r}
#No normalization
pca_gbm_norm <- prcomp(t_gbm_exp_clean[, !(names(t_gbm_exp_clean) %in% "barcode")])
pca_gbm_norm_scores <- as.data.frame(pca_gbm_norm$x)
pca_gbm_norm_scores$barcode = rownames(pca_gbm_norm_scores)
plot_pca_gbm_norm_df <- merge(pca_gbm_norm_scores, gbm_anot_clean[, c("paper_Transcriptome.Subtype", "paper_IDH.status", "barcode")], by = "barcode")
```


quantile normalisation
```{r}
#Remove non numerical columns
  expr_mat <- t_gbm_exp_clean[, !(names(t_gbm_exp_clean) %in% 
    c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status"))]
```
```{r}
#sort to obtain ranks
sorted_mat <- apply(expr_mat, 2, sort)
```
```{r}
#obtain mean
rank_means <- rowMeans(sorted_mat)
```
```{r}
#replace values by ranks
qn_mat <- expr_mat  # placeholder

for (i in 1:ncol(expr_mat)) {
  ranks <- order(expr_mat[, i])
  qn_mat[ranks, i] <- rank_means
}
```
```{r}
#plot
pca_qn <- prcomp(qn_mat)

pca_scores <- as.data.frame(pca_qn$x)
pca_scores$barcode <- rownames(pca_scores)

annot <- gbm_anot_clean[, c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status")]

plot_qn_df <- merge(pca_scores, annot, by = "barcode")
```

TMM normalisation (edgeR)
```{r}
expr_counts <- 2^(t_gbm_exp_clean[, !(names(t_gbm_exp_clean) %in% 
  c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status"))]) 

expr_counts <- as.matrix(expr_counts)
mode(expr_counts) <- "numeric"
```
```{r}

dge_obj <- DGEList(counts = expr_counts)
```
```{r}
dge_obj <- calcNormFactors(dge_obj, method = "TMM")
```
```{r}
expr_tmm_logcpm <- cpm(dge_obj, log = TRUE)
```
```{r}
pca_tmm_result <- prcomp(expr_tmm_logcpm)

pca_tmm_scores <- as.data.frame(pca_tmm_result$x)
pca_tmm_scores$barcode <- rownames(pca_tmm_scores)
```
```{r}
annotation_df <- gbm_anot_clean[, c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status")]

plot_tmm_df <- merge(pca_tmm_scores, annotation_df, by = "barcode")
```


```{r}


p1 <- ggplot(plot_tmm_df, aes(PC1, PC2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA after TMM normalization")

p2 <- ggplot(plot_qn_df, aes(PC1, PC2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA after manual quantile normalization")

p3 <- ggplot(plot_pca_gbm_norm_df, 
             aes(x = PC1, y = PC2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA without normalisation", color = "Subtype")

(p1) /
(p2) /
(p3)
```

# Exercise 4:
##Top DEGs between subtypes (one-vs-rest Wilcoxon, BH-corrected). Focus on CL vs. ME and PN vs. NE contrasts.
```{r}
# ----------------------------
# 1. Pool DEGs across subtypes
# ----------------------------
all_deg <- do.call(rbind, results_list)

# Rank by strongest effect size
all_deg <- all_deg[order(-abs(all_deg$logFC)), ]
top50_genes <- head(all_deg$gene, 50)
row.names(annot_t_gbm_exp_clean) <- annot_t_gbm_exp_clean$barcode
row.names(gbm_anot_clean) <- gbm_anot_clean$barcode

# ----------------------------
# 2. Expression matrix
# ----------------------------
expr_mat <- annot_t_gbm_exp_clean[, top50_genes]

# Scale per gene (important for heatmap patterns)
expr_mat_scaled <- scale(expr_mat)

# ----------------------------
# 3. Annotation data
# ----------------------------
ann <- gbm_anot_clean[rownames(expr_mat_scaled), c(
  "paper_Transcriptome.Subtype",
  "paper_IDH.status",
  "paper_MGMT.promoter.status",
  "paper_Grade"
)]


# ----------------------------
# 4. Handle NA values
# ----------------------------
ann$paper_Transcriptome.Subtype[is.na(ann$paper_Transcriptome.Subtype)] <- "Unknown"
ann$paper_IDH.status[is.na(ann$paper_IDH.status)] <- "Unknown"
ann$paper_MGMT.promoter.status[is.na(ann$paper_MGMT.promoter.status)] <- "Unknown"
ann$paper_Grade[is.na(ann$paper_Grade)] <- "Unknown"

row.names(ann) <- gbm_anot_clean$barcode
# Convert to factors
#ann$paper_Transcriptome.Subtype <- as.factor(ann$paper_Transcriptome.Subtype)
#ann$paper_IDH.status <- as.factor(ann$paper_IDH.status)
#ann$paper_MGMT.promoter.status <- as.factor(ann$paper_MGMT.promoter.status)
#ann$paper_Grade <- as.factor(ann$paper_Grade)

# ----------------------------
# 5. Optional: annotation colors
# ----------------------------
ann_colors <- list(
  paper_Transcriptome.Subtype = c(CL="red", ME="blue", NE="green", PN="purple", Unknown="grey"),
  paper_IDH.status = c(Mutant="orange", WT="grey", Unknown="black"),
  paper_MGMT.promoter.status = c(Methylated="darkgreen", Unmethylated="brown", Unknown="black"),
  paper_Grade = c(G4="red", Unknown="black")
)
```

```{r}
pheatmap(
  expr_mat,
  annotation_col = ann$paper_IDH.status,
  annotation_colors = ann_colors,
  show_rownames = FALSE,
  show_colnames = TRUE,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  scale = "none"
)


```




