---
title: "3_4_expressiondist_differentialexp"
output: html_document
---
```{r}
#load packages
library(dplyr)
library(ggplot2)
library(here)
library(edgeR)
library(patchwork)
library(pheatmap)
```
Load the data
```{r}
load(here("data/GBM_data.RData"))
```
Data cleaning
```{r}
# Remove gene column and set rownames to gene names
t_gbm_exp_clean <- gbm_exp_clean
t_gbm_exp_clean <- t_gbm_exp_clean[,-1]
t_gbm_exp_clean <- t(t_gbm_exp_clean)
colnames(t_gbm_exp_clean) <- gbm_exp_clean$gene
```
# Exercise 3:
## Are logCPM values approximately normal? Compare Q-Q plots across subtypes.
```{r}
t_gbm_exp_clean <- as.data.frame(t_gbm_exp_clean)
t_gbm_exp_clean$barcode <- rownames(t_gbm_exp_clean)

annot_t_gbm_exp_clean <- merge(t_gbm_exp_clean, gbm_anot_clean[, c("paper_Transcriptome.Subtype", "paper_IDH.status", "barcode","paper_Grade","paper_MGMT.promoter.status")], by = "barcode")
rownames(annot_t_gbm_exp_clean) <- annot_t_gbm_exp_clean$barcode
```
Compute QQ-Plot on distribution of expression values within samples, grouped by subtype
```{r}
expr <- annot_t_gbm_exp_clean[, !(names(annot_t_gbm_exp_clean) %in% 
  c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status"))]

subtypes <- annot_t_gbm_exp_clean$paper_Transcriptome.Subtype

values <- as.numeric(as.matrix(expr))
Subtype <- rep(subtypes, each = ncol(expr))  

plot_df <- data.frame(value = values, Subtype = Subtype)


```
## plot side by side (takes a while to load)!
```{r}
ggplot(plot_df, aes(sample = value)) +
  stat_qq(size = 0.5) +
  stat_qq_line(color = "red") +
  facet_wrap(~ Subtype) +
  theme_minimal()
```
## CPM values appear to be roughly normaly distributed
##Question 2:Does normalization method (quantile vs. TMM vs. none) affect the separation observed in PCA?
```{r}
#No normalization
pca_gbm_norm <- prcomp(t_gbm_exp_clean[, !(names(t_gbm_exp_clean) %in% "barcode")])
pca_gbm_norm_scores <- as.data.frame(pca_gbm_norm$x)
pca_gbm_norm_scores$barcode = rownames(pca_gbm_norm_scores)
plot_pca_gbm_norm_df <- merge(pca_gbm_norm_scores, gbm_anot_clean[, c("paper_Transcriptome.Subtype", "paper_IDH.status", "barcode")], by = "barcode")
```


quantile normalisation
```{r}
#Remove non numerical columns
  expr_mat <- t_gbm_exp_clean[, !(names(t_gbm_exp_clean) %in% 
    c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status"))]
```
```{r}
#sort to obtain ranks
sorted_mat <- apply(expr_mat, 2, sort)
```
```{r}
#obtain mean
rank_means <- rowMeans(sorted_mat)
```
```{r}
#replace values by ranks
qn_mat <- expr_mat  # placeholder

for (i in 1:ncol(expr_mat)) {
  ranks <- order(expr_mat[, i])
  qn_mat[ranks, i] <- rank_means
}
```
```{r}
#plot
pca_qn <- prcomp(qn_mat)

pca_scores <- as.data.frame(pca_qn$x)
pca_scores$barcode <- rownames(pca_scores)

annot <- gbm_anot_clean[, c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status")]

plot_qn_df <- merge(pca_scores, annot, by = "barcode")
```

TMM normalisation (edgeR)
```{r}
expr_counts <- 2^(t_gbm_exp_clean[, !(names(t_gbm_exp_clean) %in% 
  c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status"))]) 

expr_counts <- as.matrix(expr_counts)
mode(expr_counts) <- "numeric"
```
```{r}

dge_obj <- DGEList(counts = expr_counts)
```
```{r}
dge_obj <- calcNormFactors(dge_obj, method = "TMM")
```
```{r}
expr_tmm_logcpm <- cpm(dge_obj, log = TRUE)
```
```{r}
pca_tmm_result <- prcomp(expr_tmm_logcpm)

pca_tmm_scores <- as.data.frame(pca_tmm_result$x)
pca_tmm_scores$barcode <- rownames(pca_tmm_scores)
```
```{r}
annotation_df <- gbm_anot_clean[, c("barcode", "paper_Transcriptome.Subtype", "paper_IDH.status")]

plot_tmm_df <- merge(pca_tmm_scores, annotation_df, by = "barcode")
```


```{r}


p1 <- ggplot(plot_tmm_df, aes(PC1, PC2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA after TMM normalization")

p2 <- ggplot(plot_qn_df, aes(PC1, PC2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA after manual quantile normalization")

p3 <- ggplot(plot_pca_gbm_norm_df, 
             aes(x = PC1, y = PC2, color = paper_Transcriptome.Subtype)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA without normalisation", color = "Subtype")

(p1) /
(p2) /
(p3)
```

# Exercise 4:
##Top DEGs between subtypes (one-vs-rest Wilcoxon, BH-corrected). Focus on CL vs. ME and PN vs. NE contrasts.
```{r}
# Identify gene columns (exclude annotation columns)
gene_cols <- setdiff(
  colnames(annot_t_gbm_exp_clean),
  c("paper_Transcriptome.Subtype", "paper_IDH.status", "barcode","paper_Grade","paper_MGMT.promoter.status","barcode")
)

results_list <- list()

for (subtype in c("CL","ME","PN","NE")) {
  
  # Define groups
  group1_idx <- annot_t_gbm_exp_clean$paper_Transcriptome.Subtype == subtype
  group2_idx <- annot_t_gbm_exp_clean$paper_Transcriptome.Subtype != subtype
  
  res_subtype <- data.frame(
    gene = gene_cols,
    subtype = subtype,
    p_value = NA,
    logFC = NA
  )
  
  for (g in seq_along(gene_cols)) {
    
    gene <- gene_cols[g]
    
    x <- annot_t_gbm_exp_clean[group1_idx, gene]
    y <- annot_t_gbm_exp_clean[group2_idx, gene]
    
    # Wilcoxon test
    wt <- wilcox.test(x, y, alternative = "two.sided")
    
    # Store results
    res_subtype$p_value[g] <- wt$p.value
    
    # Optional: log fold change (median-based)
    res_subtype$logFC[g] <- median(x, na.rm = TRUE) - median(y, na.rm = TRUE)
  }
  
  # Adjust p-values (important!)
  res_subtype$padj <- p.adjust(res_subtype$p_value, method = "BH")
  
  results_list[[subtype]] <- res_subtype
}

# Combine all results
final_results <- do.call(rbind, results_list)
```
```{r}
library(dplyr)

top50 <- final_results %>%
  group_by(subtype) %>%
  arrange(desc(logFC), .by_group = TRUE) %>%
  slice_head(n = 50) %>%
  mutate(rank = row_number()) %>%
  ungroup()

top50_wide <- top50 %>%
  select(subtype, rank, gene) %>%
  tidyr::pivot_wider(names_from = subtype, values_from = gene) %>%
  arrange(rank)
```


```{r}
# -----------------------------
# 1. Top 50 genes (global DEGs)
# -----------------------------
top50_genes <- final_results %>%
  arrange(padj, desc(abs(logFC))) %>%
  distinct(gene, .keep_all = TRUE) %>%
  slice_head(n = 50) %>%
  pull(gene)

# -----------------------------
# 2. Expression matrix
# -----------------------------
expr_mat <- annot_t_gbm_exp_clean[, top50_genes]
expr_mat <- t(as.matrix(expr_mat))
colnames(expr_mat) <- annot_t_gbm_exp_clean$barcode

# -----------------------------
# 3. Annotation (CLEAN + SAFE)
# -----------------------------
annotation_col <- annot_t_gbm_exp_clean[, c(
  "paper_Transcriptome.Subtype",
  "paper_IDH.status",
  "paper_MGMT.promoter.status",
  "paper_Grade"
)]
annotation_col <- data.frame(lapply(annotation_col, function(x) {
  x <- trimws(as.character(x))
  x[is.na(x)] <- "Unknown"
  factor(x)
}))
rownames(annotation_col) <- annot_t_gbm_exp_clean$barcode

# -----------------------------
# 4. Align samples
# -----------------------------
common_samples <- intersect(colnames(expr_mat), rownames(annotation_col))
expr_mat       <- expr_mat[, common_samples]
annotation_col <- annotation_col[common_samples, ]

# -----------------------------
# 5. Scaling (gene-wise = row-wise)
# -----------------------------
expr_mat_scaled <- t(scale(t(expr_mat)))
expr_mat_scaled[!is.finite(expr_mat_scaled)] <- 0

# -----------------------------
# 6. Annotation colors
# -----------------------------
ann_colors <- list(
  paper_Transcriptome.Subtype    = c(CL="red", ME="blue", PN="green", NE="purple"),
  paper_IDH.status               = c(WT="black", Mutant="orange", Unknown="grey"),
  paper_MGMT.promoter.status     = c(Methylated="darkgreen", Unmethylated="red", Unknown="grey"),
  paper_Grade                    = c(G4="brown", Unknown="grey")
)

# -----------------------------
# 7. Sanity check
# -----------------------------
stopifnot(all(colnames(expr_mat_scaled) == rownames(annotation_col)))

# -----------------------------
# 8. Heatmap
# -----------------------------
pheatmap(
  expr_mat_scaled,
  annotation_col     = annotation_col,
  annotation_colors  = ann_colors,
  show_rownames      = FALSE,
  show_colnames      = FALSE,
  clustering_method  = "complete",
  scale              = "none"
)
```
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

# -----------------------------
# 1. Landmark genes
# -----------------------------
genes <- c("EGFR", "CHI3L1", "MET", "PDGFRA", "IDH1", "NEFL", "SYT1")

# -----------------------------
# 2. Reshape to long format
# -----------------------------
df_long <- annot_t_gbm_exp_clean %>%
  select(all_of(c("paper_Transcriptome.Subtype", genes))) %>%
  pivot_longer(
    cols = all_of(genes),
    names_to = "gene",
    values_to = "expression"
  )

# -----------------------------
# 3. Clean subtype factor order
# -----------------------------
df_long$paper_Transcriptome.Subtype <- factor(
  df_long$paper_Transcriptome.Subtype,
  levels = c("CL", "ME", "PN", "NE")
)

# -----------------------------
# 4. Boxplot (faceted by gene)
# -----------------------------
ggplot(df_long, aes(
  x = paper_Transcriptome.Subtype,
  y = expression,
  fill = paper_Transcriptome.Subtype
)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ gene, scales = "free_y") +
  theme_bw() +
  labs(
    x = "Subtype",
    y = "Expression"
  ) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```




