---
title: "5_predictivemodel"
output: html_document
---
#Excersize 5
##Target: paper_Transcriptome.Subtype (4-class). Train RF and Multinomial Logistic Regression.

In this part we want to train a random forest and multinomial logistic regression to predict the tumor subtype (paper_Transcriptome.Subtype) which can take 4 values. 

At first we need to load the data here
```{r}
library(here)
library(ggplot2)
library(patchwork)
library(dplyr)

load(here("data/GBM_data.RData"))
```
Next we will create a data frame with the patients as rows and genes + subtype as columns without any missing values (NA)
```{r}
# set rows to patient barcodes and columns to genes
data_model <- gbm_exp_clean
data_model <- data_model[,-1]
data_model <- t(data_model)
colnames(data_model) <- gbm_exp_clean$gene
data_model = scale(data_model)
data_model = as.data.frame(data_model)
# in order to join the annotation we match the rownames of data_model and the barcodes in the annotation data 
data_model_anot = data_model
idx <- match(rownames(data_model_anot), gbm_anot_clean$barcode)
data_model_anot$paper_Transcriptome.Subtype <- gbm_anot_clean$paper_Transcriptome.Subtype[idx]
# remove NAs
data_model_anot <- data_model_anot %>% na.omit()
```
Now we want to train a random forest and a multinomial logistic regression model. 

```{r}
# first we convert the subtype into a factor
data_model_anot$paper_Transcriptome.Subtype <- as.factor(data_model_anot$paper_Transcriptome.Subtype)
```
```{r}
# load some important packages
library(mlbench)
library(caret)
library(tidyverse)
library(randomForest)
library(glmnet)
```
```{r}
# Now we split the data into training and testing data sets
set.seed(123)
train_idx <- createDataPartition(data_model_anot$paper_Transcriptome.Subtype, p = 0.7, list = FALSE) # 70% of samples for training
train <- data_model_anot[train_idx, ]
test <- data_model_anot[-train_idx, ]
```
First we will train the logistic regression
```{r}
# at first separate the predictors from the subtypes as required for the input of glmnet
train_predictors <- as.matrix(train[,1:2994])
train_subtypes <- train$paper_Transcriptome.Subtype
# now we want to find the optimal lambda for our lasso regularization via cross validation using cv.glmnet
logistic_model_cv <- cv.glmnet(train_predictors, train_subtypes, family = "multinomial", alpha = 1)
plot(logistic_model_cv)
# the final model will use the optimal lambda found previously to predict the subtypes
logistic_model <- glmnet(
  train_predictors, train_subtypes, family = "multinomial", lambda = logistic_model_cv$lambda.min
)
# now we can test the model on our test data
test_predictors <- as.matrix(test[,1:2994])
test_subtypes <- test$paper_Transcriptome.Subtype

pred <- predict(logistic_model, test_predictors, type = "class")
pred <- as.factor(pred)
```
We also want to report a confusion matrix for the logistic regression now 
```{r}
confusionMatrix(data = pred, reference = test_subtypes)
```
It seems that NE is the hardest to classify. 

Next we will take a look at the feature importance, i.e. the coefficients from the MLR. 
```{r}
coef_matrix <- coef(logistic_model)
coef_subtype_CL <- coef_matrix$CL
coef_subtype_ME <- coef_matrix$ME
coef_subtype_NE <- coef_matrix$NE
coef_subtype_PN <- coef_matrix$PN

important_genes_CL <- data.frame(
  gene = rownames(coef_subtype_CL),
  coef = as.numeric(coef_subtype_CL))

# remove intercept and order according to absolute wheight
important_genes_CL <- important_genes_CL[2:2995,]
important_genes_CL <- important_genes_CL[order(abs(important_genes_CL$coef), decreasing = TRUE),]

important_genes_ME <- data.frame(
  gene = rownames(coef_subtype_ME),
  coef = as.numeric(coef_subtype_ME))

# remove intercept and order according to absolute wheight
important_genes_ME <- important_genes_ME[2:2995,]
important_genes_ME <- important_genes_ME[order(abs(important_genes_ME$coef), decreasing = TRUE),]

important_genes_NE <- data.frame(
  gene = rownames(coef_subtype_NE),
  coef = as.numeric(coef_subtype_NE))

# remove intercept and order according to absolute wheight
important_genes_NE <- important_genes_NE[2:2995,]
important_genes_NE <- important_genes_NE[order(abs(important_genes_NE$coef), decreasing = TRUE),]

important_genes_PN <- data.frame(
  gene = rownames(coef_subtype_PN),
  coef = as.numeric(coef_subtype_PN))

# remove intercept and order according to absolute wheight
important_genes_PN <- important_genes_PN[2:2995,]
important_genes_PN <- important_genes_PN[order(abs(important_genes_PN$coef), decreasing = TRUE),]
```
Now we also want to visualize the genes for each subtype
```{r}
#store top 10 genes per subtype
top_CL <- head(important_genes_CL, 10)
top_ME <- head(important_genes_ME, 10)
top_NE <- head(important_genes_NE, 10)
top_PN <- head(important_genes_PN, 10)
#plot top 10 genes per subtype 
p_CL <- ggplot(top_CL, aes(x = reorder(gene, coef), y = coef)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "CL", x = "Gene", y = "Coefficient")

p_ME <- ggplot(top_ME, aes(x = reorder(gene, coef), y = coef)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "ME", x = "Gene", y = "Coefficient")

p_NE <- ggplot(top_NE, aes(x = reorder(gene, coef), y = coef)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "NE", x = "Gene", y = "Coefficient")

p_PN <- ggplot(top_PN, aes(x = reorder(gene, coef), y = coef)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "PN", x = "Gene", y = "Coefficient")
```
```{r}
# plot all plots in one figure 
library(patchwork)
(p_CL | p_ME) / (p_NE | p_PN)
```
Next to the logistic regression we also want to train a random forest model
```{r}
# we will use the train_predictors and train_subtypes to train the model. We have defined them alreacy for the logistic regression model
rf_model <- randomForest(x = train_predictors, y = train_subtypes, ntree = 500, importance = TRUE, mtry = floor(sqrt(ncol(train_predictors))))
```
```{r}
# predict on the test data
pred_rf <- predict(rf_model, test_predictors)
```
```{r}
# compute the confusion matrix 
confusionMatrix(pred_rf, test_subtypes)
```
Also here the NE subtype seems to be the hardest to classify 

Since there are no coefficients in RandomForest we can only extract the importance values
```{r}
importance_vals <- importance(rf_model)

importance_df <- data.frame(
  gene = rownames(importance_vals),
  importance = importance_vals[, 1])

importance_df <- importance_df[
  order(importance_df$importance, decreasing = TRUE),]
```
Hence we will also plot the top important genes for the random forest model
```{r}
top_genes_rf <- head(importance_df, 20)

ggplot(top_genes_rf, aes(x = reorder(gene, importance), y = importance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top Important Genes (Random Forest)",
       x = "Gene", y = "Importance")
```
# Excersize 7 
## Compare RF vs. MLR performance (accuracy, F1, per-class precision/recall).
```{r}
# first we define the predictions for each model again
pred_rf = as.factor(pred_rf)
pred_mlr = as.factor(pred)
# we also define the actual subtypes again as factors 
y_test <- as.factor(test_subtypes)
# Now we can compute the confusion matrices
cm_rf  <- confusionMatrix(pred_rf, y_test)
cm_mlr <- confusionMatrix(pred_mlr, y_test)
# From here we can compute the accuracy of both models 
acc_rf  <- cm_rf$overall["Accuracy"]
acc_mlr <- cm_mlr$overall["Accuracy"]
# We can also compute the per class precision and recall
rf_class <- data.frame(
  Class = rownames(cm_rf$byClass),
  Precision = cm_rf$byClass[, "Pos Pred Value"],
  Recall    = cm_rf$byClass[, "Sensitivity"])

mlr_class <- data.frame(
  Class = rownames(cm_mlr$byClass),
  Precision = cm_mlr$byClass[, "Pos Pred Value"],
  Recall    = cm_mlr$byClass[, "Sensitivity"])
```
```{r}
# In order to compute the F1 score we need the following function
f1_score <- function(precision, recall) {(2 * precision * recall) / (precision + recall)}
# Now we can calculate the F1 values
rf_class$F1  <- f1_score(rf_class$Precision, rf_class$Recall)
mlr_class$F1 <- f1_score(mlr_class$Precision, mlr_class$Recall)
# To compute the F1 score per model we can average the F1 score wheighted respectively to the proportions of counts
class_counts = as.numeric(table(y_test))

weighted_f1_rf <- sum(rf_class$F1 * class_counts) / sum(class_counts)
weighted_f1_mlr <- sum(mlr_class$F1 * class_counts) / sum(class_counts)
```
Now we can summarize the results in a single dataframe
```{r}
results_models <- data.frame(
  Model = c("Random Forest", "Multinomial Logistic"),
  Accuracy = c(acc_rf, acc_mlr),
  Weighted_F1 = c(weighted_f1_rf, weighted_f1_mlr)
)
results_models
```
We can also compare the results for the different classes in each model
```{r}
rf_class$model <- "RF"
mlr_class$model <- "MLR"

comparison_classes <- rbind(rf_class, mlr_class)
comparison_classes
```
This data can also be vizualized by plotting the F1 scores for each subtype
```{r}
ggplot(comparison_classes, aes(x = Class, y = F1, fill = model)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Per-class F1 comparison", y = "F1 score")
```
We can also plot the precision and recall for each class in each model in order to see which model performs the best (precision = TP / pred_positives, recall = TP / positives)
```{r}
ggplot(comparison_classes, aes(x = Recall, y = Precision, color = model)) +
  geom_point(size = 3) +
  theme_minimal() +
  facet_wrap(~Class) +
  labs(title = "Precision vs Recall by Class")
```
## Are predictive genes also DEGs? Venn diagram or UpSet plot.

## Biological interpretation: which pathways are enriched in the top 10 predictive genes?
To answer this question we will take the top 10 predictive genes for each subtype and perform an enrichment analysis 

In order to answer this question we need to run an enrichment analysis on the top 10 predicitive genes for each subtype
```{r}
# load important packages 
install.packages("BiocManager")
BiocManager::install("clusterProfiler")
BiocManager::install("org.Hs.eg.db")
```
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
```

GO analysis returns enriched biological pathways -> did not wiokrt well
```{r}
# first we extract the gene names of the top ten predicitive genes of the CL subtype
genes_CL <- head(important_genes_CL$gene, 10)
# next we convert the gene names into ids using the org.HS.eg.db data base 
genes_CL_entrez <- bitr(genes_CL, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db
)
# Next we can run the gene ontology enrichment analysis 
ego_CL <- enrichGO(
  gene = genes_CL_entrez$ENTREZID,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          # Biological Process
  pAdjustMethod = "BH",
  readable = TRUE
)
# visualize the results
dotplot(ego_CL, showCategory = 10)
```
```{r}
# first we extract the gene names of the top ten predicitive genes of the ME subtype
genes_ME <- head(important_genes_ME$gene, 10)
# next we convert the gene names into ids using the org.HS.eg.db data base 
genes_ME_entrez <- bitr(genes_ME, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db
)
# Next we can run the gene ontology enrichment analysis 
ego_ME <- enrichGO(
  gene = genes_ME_entrez$ENTREZID,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          # Biological Process
  pAdjustMethod = "BH",
  readable = TRUE
)
# visualize the results
dotplot(ego_ME, showCategory = 10)
```
```{r}
# first we extract the gene names of the top ten predicitive genes of the NE subtype
genes_NE <- head(important_genes_NE$gene, 10)
# next we convert the gene names into ids using the org.HS.eg.db data base 
genes_NE_entrez <- bitr(genes_NE, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db
)
# Next we can run the gene ontology enrichment analysis 
ego_NE <- enrichGO(
  gene = genes_NE_entrez$ENTREZID,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          # Biological Process
  pAdjustMethod = "BH",
  readable = TRUE
)
# visualize the results
dotplot(ego_NE, showCategory = 10)
```
```{r}
# first we extract the gene names of the top ten predicitive genes of the PN subtype
genes_PN <- head(important_genes_PN$gene, 10)
# next we convert the gene names into ids using the org.HS.eg.db data base 
genes_PN_entrez <- bitr(genes_PN, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db
)
# Next we can run the gene ontology enrichment analysis 
ego_PN <- enrichGO(
  gene = genes_PN_entrez$ENTREZID,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          # Biological Process
  pAdjustMethod = "BH",
  readable = TRUE
)
# visualize the results
dotplot(ego_PN, showCategory = 10)
```

Since GO did not work correctly we decided to run GSEA as well 
```{r}
BiocManager::install("enrichplot")
BiocManager::install("msigdbr")
```
Next we need to bring the genes into a ranked list format based on their coefficient
```{r}
ranked_CL_genes <- important_genes_CL %>% arrange(desc(abs(coef)))
gene_list_CL <- ranked_CL_genes$coef
names(gene_list_CL) <- ranked_CL_genes$gene
# now convert gene names to entrez ids again 
gene_df_CL <- bitr(
  names(gene_list_CL),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)
# now rename the ranked list using the entrez ids
gene_list_CL <- gene_list_CL[gene_df_CL$SYMBOL]
names(gene_list_CL) <- gene_df_CL$ENTREZID

gene_list_CL = sort(gene_list_CL, decreasing = T)
# this way we can now compute the gsea
library(msigdbr)

hallmark <- msigdbr(species = "Homo sapiens", category = "H")

hallmark_df <- hallmark[, c("gs_name", "gene_symbol")]

gsea_CL_hallmark <- GSEA(
  geneList = gene_list_CL,
  TERM2GENE = hallmark_list,
  pvalueCutoff = 0.05
)
```




