---
title: "A_B_cleaning_exploration"
output: html_document
---
# Annotation cleaning
## Tidy the annotation file to keep only curated, paper-level columns.
```{r}
# at first we only select the paper-level columns
anot_clean <- gbm_anot_clean[, c("barcode", grep("^paper_", names(gbm_anot_clean),
value = TRUE))]
```
This way we keep 52 annotation columns containing information on whether certain analysis has been performed, where the tissue is from, but also numerical values like age or mutation count. 

# Clinical Variable Exploration
## Explore paper_IDH.status, paper_MGMT.promoter.status, paper_Grade, paper_Age..years.at.diagnosis
```{r}
summary(anot_clean[c("paper_IDH.status", "paper_MGMT.promoter.status", "paper_Grade", "paper_Age..years.at.diagnosis.")])
```
## Tabulate paper_Karnofsky.Performance.Score (functional status) and paper_Mutation.Count.
```{r}
table(anot_clean$paper_Karnofsky.Performance.Score)
table(anot_clean$paper_Mutation.Count)
```
## Investigate paper_Telomere.Maintenance (ALT vs. TERT) — does it associate with subtype?
```{r}
# create a count table counting telomere Mainrtainance type in each subtype
tab_telomere_subtype <- table(anot_clean$paper_Telomere.Maintenance, anot_clean$paper_Transcriptome.Subtype)
tab_telomere_subtype
# testing association in fisher exact test
test_telomere_subtype <- fisher.test(tab_telomere_subtype)
test_telomere_subtype
```
ATRX is strongly associated with PN, TERT seems associated with CL and ME equally. The Fisher Exact Test tells us that there is a significant association between Telomere.Maintenance and Subtype


