---
title: "C_featureclassification"
output: html_document
---
# Step C: Clinical Feature Classification -- Predict a binary or multi-class clinical outcome using clinical variables only
```{r}
# at first we only select the paper-level columns
anot_clean <- gbm_anot_clean[, c("barcode", grep("^paper_", names(gbm_anot_clean),
value = TRUE))]
```
## Define the outcome variable and apply appropriate encoding.

We have decided to predict the clinical variable paper_IDH.status (Isocitrate Dehydrogenase) which is binary and can either be WT or mutant. IDH mutations are generally associated with better disease-outcome making it an interesting variable to predict.
```{r}
# At first lets examine the paper_IDH.status column
summary(anot_clean$paper_IDH.status)
table(anot_clean$paper_IDH.status)
sum(is.na(anot_clean$paper_IDH.status))
```
This shows that for 305 out of 325 there is a value in the respective column. Out of these 305, 285 patients have the WT IDH status and 20 exhibit a mutant version. 

## Select a small set of meaningful features and handle missing values.
```{r}
# im ersten Schritt definieren wir unsere target variable und einige scheinbar sinnvolle features, welche nicht direkt mit dem IDH status zusammenhängen -> leakage vermeiden
target <- "paper_IDH.status"
features <- c(
  "paper_Age..years.at.diagnosis.",
  "paper_Gender",
  "paper_Histology",
  "paper_Grade",
  "paper_Karnofsky.Performance.Score",
  "paper_Mutation.Count",
  "paper_Percent.aneuploidy",
  "paper_MGMT.promoter.status",
  "paper_Chr.7.gain.Chr.10.loss",
  "paper_Chr.19.20.co.gain",
  "paper_TERT.promoter.status",
  "paper_TERT.expression..log2.",
  "paper_ATRX.status",
  "paper_BRAF.V600E.status",
  "paper_ABSOLUTE.purity",
  "paper_ABSOLUTE.ploidy",
  "paper_ESTIMATE.stromal.score",
  "paper_ESTIMATE.immune.score"
)
# then we can define a new data frame containing only the variables in the features and the target
anot_clean_IDH <- anot_clean %>%
  dplyr::select(all_of(c(target, features))) %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(paper_IDH.status = as.factor(paper_IDH.status))
```
```{r}
# next we can take a look at the missing values in each variable
library(naniar)
miss_var_summary(anot_clean_IDH)
vis_miss(anot_clean_IDH)
```
Since paper_TERT.promoter.status has 85% missing values we will exclude it from further analysis and imputation
```{r}
anot_clean_IDH2 <- anot_clean_IDH[which(names(anot_clean_IDH) != "paper_TERT.promoter.sta")]
```
For the other variables we will perform imputations, meaning that missing values will be replaced by computationally determinded values.
```{r}
# imputation with mice package chooses appropriate imputation strategy for each variable (numerical, logreg, polyreg, polr)
library(mice)
imp <- mice(anot_clean_IDH2, m = 1, method = NULL, seed = 42, printFlag = FALSE)
# now we can extract the complete data set with no missing values
anot_clean_IDH_complete <- complete(imp, 1)
miss_var_summary(anot_clean_IDH_complete)
```
We will exclude paper Histology and paper Grade since they only take one value which also causes them to fail at the imputation.
```{r}
anot_clean_IDH_complete <- anot_clean_IDH_complete %>% dplyr::select(-c(paper_Histology, paper_Grade))
```
We now have a dataframe with no missing values. 
##  Train/test split (70/30 or 5-fold CV). Fit: Logistic Regression, Random Forest, k-NN.
```{r}
library(caret)

# first we need to split the data into a testing and training dataset
set.seed(42)
train_index <- createDataPartition(anot_clean_IDH_complete$paper_IDH.status, 
                                   p = 0.7, list = FALSE)
train_data_IDH <- anot_clean_IDH_complete[train_index, ]
test_data_IDH  <- anot_clean_IDH_complete[-train_index, ]

# alternatively we can also do a 5-fold crossvalidation
ctrl <- trainControl(method = "cv", number = 5,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary)

# first fit the logistic regression
set.seed(42)
model_lr <- caret::train(paper_IDH.status ~ ., data = train_data_IDH, method = "glmnet", family = "binomial", trControl = ctrl, metric = "ROC",tuneGrid = expand.grid(alpha = 1,lambda = seq(0.0001, 1, length = 50))) # with lasso regularization

# next fit the random forest model
set.seed(42)
model_rf <- caret::train(paper_IDH.status ~ ., data = train_data_IDH,
                  method = "rf",
                  trControl = ctrl, metric = "ROC", 
                  tuneLength = 10) # tunes mtry

# lastly fit the knn model
set.seed(42)
model_knn <- caret::train(paper_IDH.status ~ ., data = train_data_IDH,
                   method = "knn",
                   preProcess = c("center", "scale"),  # wichtig für kNN!
                   trControl = ctrl, metric = "ROC", 
                   tuneLength = 10) # tunes k
```
## Report confusion matrices, accuracy, sensitivity/specificity, and AUC-ROC.
```{r}
# at first we predict the IDH statuses of the test data patients
pred_lr  <- predict(model_rf,  test_data_IDH)
pred_rf  <- predict(model_rf,  test_data_IDH)
pred_knn <- predict(model_knn, test_data_IDH)

# for the ROC curve we need the probabilities for the predicted statuses as well
prob_lr  <- predict(model_lr,  test_data_IDH, type = "prob")[, "Mutant"]
prob_rf  <- predict(model_rf,  test_data_IDH, type = "prob")[, "Mutant"]
prob_knn <- predict(model_knn, test_data_IDH, type = "prob")[, "Mutant"]
```
Next we can compute the confusion matrices 
```{r}
cm_lr  <- confusionMatrix(pred_lr,  test_data_IDH$paper_IDH.status, positive = "Mutant")
cm_rf  <- confusionMatrix(pred_rf,  test_data_IDH$paper_IDH.status, positive = "Mutant")
cm_knn <- confusionMatrix(pred_knn, test_data_IDH$paper_IDH.status, positive = "Mutant")

cm_lr
cm_rf
cm_knn
```
Next we can also compute the AUC of the ROC curves
```{r}
# first we install a package to compute the ROC curves
library(pROC)
# we compute the roc curves for each model
roc_lr  <- roc(test_data_IDH$paper_IDH.status, prob_lr,  levels = c("WT", "Mutant"))
roc_rf  <- roc(test_data_IDH$paper_IDH.status, prob_rf,  levels = c("WT", "Mutant"))
roc_knn <- roc(test_data_IDH$paper_IDH.status, prob_knn, levels = c("WT", "Mutant"))
# from the roc curves we can also extract the aucs
auc(roc_lr)
auc(roc_rf)
auc(roc_knn)
# additionally we can plot the ROC curves and the respective AUCs in a single plot
plot(roc_lr,  col = "blue",  main = "ROC Curves")
plot(roc_rf,  col = "red",   add = TRUE)
plot(roc_knn, col = "green", add = TRUE)
legend("bottomright", 
       legend = c(paste("LR  AUC =", round(auc(roc_lr), 3)),
                  paste("RF  AUC =", round(auc(roc_rf), 3)),
                  paste("kNN AUC =", round(auc(roc_knn), 3))),
       col = c("blue", "red", "green"), lwd = 2)
```
Lastly we can combine all the results in a single dataframe
```{r}
results_summary <- tibble(
  Model       = c("Logistic Regression", "Random Forest", "k-NN"),
  Accuracy    = c(cm_lr$overall["Accuracy"],
                  cm_rf$overall["Accuracy"],
                  cm_knn$overall["Accuracy"]),
  Sensitivity = c(cm_lr$byClass["Sensitivity"],
                  cm_rf$byClass["Sensitivity"],
                  cm_knn$byClass["Sensitivity"]),
  Specificity = c(cm_lr$byClass["Specificity"],
                  cm_rf$byClass["Specificity"],
                  cm_knn$byClass["Specificity"]),
  AUC         = c(as.numeric(auc(roc_lr)),
                  as.numeric(auc(roc_rf)),
                  as.numeric(auc(roc_knn)))
) %>% mutate(across(where(is.numeric), ~ round(., 3)))

print(results_summary)
```
## Feature importance from RF (variable importance) and Logistic Regression (coefficients).
At first we will take a look at the variable importances from the RF
```{r}
library(ggplot2)

# extract the variable importances from the rf model
rf_importance <- varImp(model_rf, scale = TRUE)$importance %>%
  rownames_to_column("Variable") %>%
  arrange(desc(Overall)) %>%
  slice_head(n = 15)  # Top 15

# Plot the top 15 important variables
ggplot(rf_importance, aes(x = reorder(Variable, Overall), y = Overall)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest – Variable Importance",
       x = NULL, y = "Importance (scaled)") +
  theme_minimal()
```
Next we will also extract the coefficients from the logistic regression
```{r}
lasso_coefs <- coef(model_lr$finalModel, s = model_lr$bestTune$lambda) %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  dplyr::rename(Coefficient = 2) %>%
  dplyr::filter(Variable != "(Intercept)", Coefficient != 0) %>%
  dplyr::arrange(desc(abs(Coefficient)))

# Plot
ggplot(lr_coefs, aes(x = reorder(Variable, Coefficient),
                      y = Coefficient,
                      fill = Coefficient > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
                    labels = c("TRUE" = "→ Mutant", "FALSE" = "→ WT")) +
  labs(title = "Logistic Regression – Coefficients",
       x = NULL, y = "Coefficient", fill = "Direction") +
  theme_minimal()
```




