---
title: "C_featureclassification"
output: html_document
---
# Step C: Clinical Feature Classification -- Predict a binary or multi-class clinical outcome using clinical variables only
```{r}
# at first we only select the paper-level columns
anot_clean <- gbm_anot_clean[, c("barcode", grep("^paper_", names(gbm_anot_clean),
value = TRUE))]
```
## Define the outcome variable and apply appropriate encoding.

We have decided to predict the clinical variable paper_IDH.status (Isocitrate Dehydrogenase) which is binary and can either be WT or mutant. IDH mutations are generally associated with better disease-outcome making it an interesting variable to predict.
```{r}
# At first lets examine the paper_IDH.status column
summary(anot_clean$paper_IDH.status)
table(anot_clean$paper_IDH.status)
sum(is.na(anot_clean$paper_IDH.status))
```
This shows that for 305 out of 325 there is a value in the respective column. Out of these 305, 285 patients have the WT IDH status and 20 exhibit a mutant version. 

## Select a small set of meaningful features and handle missing values.
```{r}
# im ersten Schritt definieren wir unsere target variable und einige scheinbar sinnvolle features, welche nicht direkt mit dem IDH status zusammenhängen -> leakage vermeiden
target <- "paper_IDH.status"
features <- c(
  "paper_Age..years.at.diagnosis.",
  "paper_Gender",
  "paper_Karnofsky.Performance.Score",
  "paper_Mutation.Count",
  "paper_Percent.aneuploidy",
  "paper_MGMT.promoter.status",
  "paper_Chr.7.gain.Chr.10.loss",
  "paper_Chr.19.20.co.gain",
  "paper_TERT.promoter.status",
  "paper_TERT.expression..log2.",
  "paper_ATRX.status",
  "paper_ABSOLUTE.purity",
  "paper_ABSOLUTE.ploidy",
  "paper_ESTIMATE.stromal.score",
  "paper_ESTIMATE.immune.score",
  "paper_BRAF.V600E.status",
  "barcode"
)
# then we can define a new data frame containing only the variables in the features and the target
anot_clean_IDH <- anot_clean %>%
  dplyr::select(all_of(c(target, features))) %>%
  column_to_rownames("barcode") %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(paper_IDH.status = as.factor(paper_IDH.status))
```
```{r}
# next we can take a look at the missing values in each variable
library(naniar)
miss_var_summary(anot_clean_IDH)
vis_miss(anot_clean_IDH)
```
Since paper_TERT.promoter.status has 85% missing values we will exclude it from further analysis and imputation
```{r}
anot_clean_IDH2 <- anot_clean_IDH %>% dplyr::select(-c(paper_TERT.promoter.status))
```
For the other variables we will perform imputations, meaning that missing values will be replaced by computationally determinded values.
```{r}
# imputation with mice package chooses appropriate imputation strategy for each variable (numerical, logreg, polyreg, polr)
library(mice)
imp <- mice(anot_clean_IDH2, m = 1, method = NULL, seed = 42, printFlag = FALSE)
# now we can extract the complete data set with no missing values
anot_clean_IDH_complete <- complete(imp, 1)
miss_var_summary(anot_clean_IDH_complete)
```
We now have a dataframe with no missing values. 
##  Train/test split (70/30 or 5-fold CV). Fit: Logistic Regression, Random Forest, k-NN.
```{r}
library(caret)

# first we need to split the data into a testing and training dataset -> since there are only very few positives (mutants) in the IDH status we decided to do a 60/40 data partition to get sufficient amount of positives in the testing data set as well
set.seed(42)
train_index <- createDataPartition(anot_clean_IDH_complete$paper_IDH.status, 
                                   p = 0.6, list = FALSE)
train_data_IDH <- anot_clean_IDH_complete[train_index, ]
test_data_IDH  <- anot_clean_IDH_complete[-train_index, ]

# alternatively we can also do a 5-fold crossvalidation
ctrl <- trainControl(method = "cv", number = 5,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary) # takes class imbalance into account

# first fit the logistic regression
set.seed(42)
model_lr <- caret::train(paper_IDH.status ~ ., data = train_data_IDH, method = "glmnet", family = "binomial", trControl = ctrl, metric = "ROC",tuneGrid = expand.grid(alpha = 1,lambda = seq(0.0001, 1, length = 50))) # with lasso regularization

# next fit the random forest model
set.seed(42)
model_rf <- caret::train(paper_IDH.status ~ ., data = train_data_IDH,
                  method = "rf",
                  trControl = ctrl, metric = "ROC", 
                  tuneLength = 10) # tunes mtry

# lastly fit the knn model
set.seed(42)
model_knn <- caret::train(paper_IDH.status ~ ., data = train_data_IDH,
                   method = "knn",
                   preProcess = c("center", "scale"),  # wichtig für kNN!
                   trControl = ctrl, metric = "ROC", 
                   tuneLength = 10) # tunes k
```
## Report confusion matrices, accuracy, sensitivity/specificity, and AUC-ROC.
```{r}
# at first we predict the IDH statuses of the test data patients
pred_lr  <- predict(model_rf,  test_data_IDH)
pred_rf  <- predict(model_rf,  test_data_IDH)
pred_knn <- predict(model_knn, test_data_IDH)

# for the ROC curve we need the probabilities for the predicted statuses as well
prob_lr  <- predict(model_lr,  test_data_IDH, type = "prob")[, "Mutant"]
prob_rf  <- predict(model_rf,  test_data_IDH, type = "prob")[, "Mutant"]
prob_knn <- predict(model_knn, test_data_IDH, type = "prob")[, "Mutant"]
```
Next we can compute the confusion matrices 
```{r}
cm_lr  <- confusionMatrix(pred_lr,  test_data_IDH$paper_IDH.status, positive = "Mutant")
cm_rf  <- confusionMatrix(pred_rf,  test_data_IDH$paper_IDH.status, positive = "Mutant")
cm_knn <- confusionMatrix(pred_knn, test_data_IDH$paper_IDH.status, positive = "Mutant")

cm_lr
cm_rf
cm_knn
```
Next we can also compute the AUC of the ROC curves
```{r}
# first we install a package to compute the ROC curves
library(pROC)
# we compute the roc curves for each model
roc_lr  <- roc(test_data_IDH$paper_IDH.status, prob_lr,  levels = c("WT", "Mutant"))
roc_rf  <- roc(test_data_IDH$paper_IDH.status, prob_rf,  levels = c("WT", "Mutant"))
roc_knn <- roc(test_data_IDH$paper_IDH.status, prob_knn, levels = c("WT", "Mutant"))
# from the roc curves we can also extract the aucs
auc(roc_lr)
auc(roc_rf)
auc(roc_knn)
# additionally we can plot the ROC curves and the respective AUCs in a single plot
plot(roc_lr,  col = "blue",  main = "ROC Curves")
plot(roc_rf,  col = "red",   add = TRUE)
plot(roc_knn, col = "green", add = TRUE)
legend("bottomright", 
       legend = c(paste("LR  AUC =", round(auc(roc_lr), 3)),
                  paste("RF  AUC =", round(auc(roc_rf), 3)),
                  paste("kNN AUC =", round(auc(roc_knn), 3))),
       col = c("blue", "red", "green"), lwd = 2)
```
Lastly we can combine all the results in a single dataframe
```{r}
results_summary <- tibble(
  Model       = c("Logistic Regression", "Random Forest", "k-NN"),
  Accuracy    = c(cm_lr$overall["Accuracy"],
                  cm_rf$overall["Accuracy"],
                  cm_knn$overall["Accuracy"]),
  Sensitivity = c(cm_lr$byClass["Sensitivity"],
                  cm_rf$byClass["Sensitivity"],
                  cm_knn$byClass["Sensitivity"]),
  Specificity = c(cm_lr$byClass["Specificity"],
                  cm_rf$byClass["Specificity"],
                  cm_knn$byClass["Specificity"]),
  AUC         = c(as.numeric(auc(roc_lr)),
                  as.numeric(auc(roc_rf)),
                  as.numeric(auc(roc_knn)))
) %>% mutate(across(where(is.numeric), ~ round(., 3)))

print(results_summary)
```
## Feature importance from RF (variable importance) and Logistic Regression (coefficients).
At first we will take a look at the variable importances from the RF
```{r}
library(ggplot2)

# extract the variable importances from the rf model
rf_importance <- varImp(model_rf, scale = TRUE)$importance %>%
  rownames_to_column("Variable") %>%
  arrange(desc(Overall)) %>%
  slice_head(n = 15)  # Top 15

# Plot the top 15 important variables
ggplot(rf_importance, aes(x = reorder(Variable, Overall), y = Overall)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest – Variable Importance",
       x = NULL, y = "Importance (scaled)") +
  theme_minimal()
```
Next we will also extract the coefficients from the logistic regression
```{r}
lasso_coefs <- coef(model_lr$finalModel, s = model_lr$bestTune$lambda) %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  dplyr::rename(Coefficient = 2) %>%
  dplyr::filter(Variable != "(Intercept)") %>%
  dplyr::arrange(desc(abs(Coefficient)))

# Plot
ggplot(lasso_coefs, aes(x = reorder(Variable, Coefficient),
                      y = Coefficient,
                      fill = Coefficient > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
                    labels = c("TRUE" = "→ Mutant", "FALSE" = "→ WT")) +
  labs(title = "Logistic Regression – Coefficients",
       x = NULL, y = "Coefficient", fill = "Direction") +
  theme_minimal()
```
# D Integrative Analysis
## Build a multimodal classifier: combine top expression features with key clinical features. Does
adding clinical data improve predictive performance?
-> For this excersize we will build a classifier based on expression data to also predict the IDH status and then we will combine it with the clinical variables we have used before to see whether this enhances the model
```{r}
# first we will perform similar data preparation steps for the expression data as we have conducted them before 
# set rows to patient barcodes and columns to genes
data_model_exp <- gbm_exp_clean
data_model_exp <- data_model_exp[,-1]
data_model_exp <- t(data_model_exp)
colnames(data_model_exp) <- gbm_exp_clean$gene
data_model_exp = scale(data_model_exp)
data_model_exp = as.data.frame(data_model_exp)
# next we will join the annotation column for the IDH status
anot_model <- anot_clean_IDH_complete %>%
  rownames_to_column("barcode") %>%
  dplyr::select(-paper_IDH.status)

data_model_exp <- data_model_exp %>%
  rownames_to_column("barcode") %>%
  left_join(anot_clean_IDH_complete %>%
              rownames_to_column("barcode") %>%    # ← rownames → Spalte
              dplyr::select(barcode, paper_IDH.status),
            by = "barcode") %>%
  column_to_rownames("barcode")
```
Now we can create a 70/30 data partition and train a logistic regression model to predict the IDH status based on the expression data
```{r}
# 70/30 Split -> since there are only very few positives (mutants) in the IDH status we decided to do a 60/40 data partition to get sufficient amount of positives in the testing data set as well
set.seed(42)
train_index <- createDataPartition(data_model_exp$paper_IDH.status, 
                                   p = 0.6, list = FALSE)
train_exp <- data_model_exp[train_index, ]
test_exp  <- data_model_exp[-train_index, ]

# create a 5 fold cross validation that takes class imbalance into account (smote)
ctrl <- trainControl(method = "cv", number = 5,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary)
# train the model using glmnet -> LASSO 
set.seed(42)
tune_grid <- expand.grid(alpha  = 1, lambda = 10^seq(-4, 0, length = 50))  # 50 lambda values

# train a logistic regression model with glmnet -> uses Lasso regularization automatically
set.seed(42)
model_exp_lr <- caret::train(paper_IDH.status ~ ., 
                             data = train_exp,
                             method = "glmnet",
                             family = "binomial",
                             trControl = ctrl,
                             metric = "ROC",
                             tuneGrid = tune_grid)

# Evaluation
pred_exp_lr  <- predict(model_exp_lr, test_exp)
prob_exp_lr  <- predict(model_exp_lr, test_exp, type = "prob")[, "Mutant"]

confusionMatrix(pred_exp_lr, test_exp$paper_IDH.status, positive = "Mutant")

roc_exp_lr <- roc(test_exp$paper_IDH.status, prob_exp_lr, levels = c("WT", "Mutant"))
auc(roc_exp_lr)
```
We can also plot all non zero coefficients
```{r}
# Koeffizienten extrahieren
exp_coefs <- coef(model_exp_lr$finalModel, model_exp_lr$bestTune$lambda) %>%
  as.matrix() %>%
  as.data.frame() %>%
  setNames("Coefficient") %>%
  rownames_to_column("Variable") %>%
  filter(Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(abs(Coefficient)))

print(exp_coefs)

# Plot
ggplot(exp_coefs, aes(x = reorder(Variable, Coefficient),
                      y = Coefficient,
                      fill = Coefficient > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
                    labels = c("TRUE" = "→ Mutant", "FALSE" = "→ WT")) +
  labs(title = "Non-zero Coefficients – Expression Model",
       x = NULL, y = "Coefficient", fill = "Direction") +
  theme_minimal()
```

Now we want to train another model including clinical data (difined variables) so see whether the performance is imporved.
```{r}
#first we prepare a combined data set with the anotation data and expression data joined by barcode and only 1 columns for IDH status
anot_model <- anot_clean_IDH_complete %>%
  dplyr::select(paper_BRAF.V600E.status,
                paper_ATRX.status) %>%
  rownames_to_column("barcode")

data_model_combined <- data_model_exp %>%
  rownames_to_column("barcode") %>%
  left_join(anot_model, by = "barcode") %>%
  column_to_rownames("barcode")
```
Now we can create a data partition again and train a model
```{r}
# create a data partition -> since there are only very few positives (mutants) in the IDH status we decided to do a 60/40 data partition to get sufficient amount of positives in the testing data set as well
set.seed(42)
train_index_comb <- createDataPartition(data_model_combined$paper_IDH.status,
                                        p = 0.6, list = FALSE)
train_comb <- data_model_combined[train_index_comb, ]
test_comb  <- data_model_combined[-train_index_comb, ]

ctrl <- trainControl(method = "cv", number = 5,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary)

# train the model using glmnet -> LASSO 
set.seed(42)
tune_grid <- expand.grid(
  alpha  = 1,                          # LASSO only
  lambda = 10^seq(-4, 0, length = 50)  # 50 lambda values
)
model_combined <- caret::train(paper_IDH.status ~ .,
                               data = train_comb,
                               method = "glmnet",
                               family = "binomial",
                               tuneGrid = tune_grid,
                               trControl = ctrl,
                               metric = "ROC")
# Evaluation
pred_comb <- predict(model_combined, test_comb)
prob_comb <- predict(model_combined, test_comb, type = "prob")[, "Mutant"]

confusionMatrix(pred_comb, test_comb$paper_IDH.status, positive = "Mutant")

roc_comb <- roc(test_comb$paper_IDH.status, prob_comb, levels = c("WT", "Mutant"))
auc(roc_comb)
```
Now lets compare the models
```{r}
# Predictions for all different models
# Expression only
pred_exp  <- predict(model_exp_lr, test_exp)
prob_exp  <- predict(model_exp_lr, test_exp, type = "prob")[, "Mutant"]

# Combined
pred_comb <- predict(model_combined, test_comb)
prob_comb <- predict(model_combined, test_comb, type = "prob")[, "Mutant"]

# Clinical only (coded above)
pred_lr   <- predict(model_lr, test_data_IDH)
prob_lr   <- predict(model_lr, test_data_IDH, type = "prob")[, "Mutant"]

# compute roc
roc_exp  <- roc(test_exp$paper_IDH.status,  prob_exp,  levels = c("WT", "Mutant"))
roc_comb <- roc(test_comb$paper_IDH.status, prob_comb, levels = c("WT", "Mutant"))
roc_lr   <- roc(test_data_IDH$paper_IDH.status, prob_lr,   levels = c("WT", "Mutant"))

# combine all results in a tibble 
results_all <- tibble(
  Model       = c("Clinical only", "Expression only", "Combined"),
  AUC         = round(c(as.numeric(auc(roc_lr)),
                        as.numeric(auc(roc_exp)),
                        as.numeric(auc(roc_comb))), 3),
  Sensitivity = round(c(
    confusionMatrix(pred_lr,   test_data_IDH$paper_IDH.status, positive = "Mutant")$byClass["Sensitivity"],
    confusionMatrix(pred_exp,  test_exp$paper_IDH.status,  positive = "Mutant")$byClass["Sensitivity"],
    confusionMatrix(pred_comb, test_comb$paper_IDH.status, positive = "Mutant")$byClass["Sensitivity"]
  ), 3),
  Specificity = round(c(
    confusionMatrix(pred_lr,   test_data_IDH$paper_IDH.status, positive = "Mutant")$byClass["Specificity"],
    confusionMatrix(pred_exp,  test_exp$paper_IDH.status,  positive = "Mutant")$byClass["Specificity"],
    confusionMatrix(pred_comb, test_comb$paper_IDH.status, positive = "Mutant")$byClass["Specificity"]
  ), 3)
)

print(results_all)

# Plot
plot(roc_lr,   col = "blue",  main = "ROC Curves")
plot(roc_exp,  col = "red",   add = TRUE)
plot(roc_comb, col = "green", add = TRUE)
legend("bottomright",
       legend = c(paste("Clinical    AUC =", round(auc(roc_lr),   3)),
                  paste("Expression  AUC =", round(auc(roc_exp),  3)),
                  paste("Combined    AUC =", round(auc(roc_comb), 3))),
       col = c("blue", "red", "green"), lwd = 2)
```




